My title

### 2023-12-01 outputs/paper/paper.qmd
Finish the drafts for Data and Methods

library(tidyverse)
library(here)
library(dplyr)
library(knitr)
```

@@

## Center for Systems Science and Engineering at JHU
The Center for Systems Science and Engineering at JHU is the center at the Department of Civils and Engineering to collect the local, national and global multidimensional data including medicine, health care, disaster response etc. During the pandemic, they collected the US and global COVID cases and deaths and report them on their GitHub. Their data is summarized by daily reports, ranging from April 12, 2020 to March 9, 2023.

To illustrate the impact of COVID on the 2020 Election to the greatest extend, the COVID data used in this paper will be the daily report on November 3, 2020 which is the Election Day [citation]. This report contains the aggregated cases and deaths for each each country and counties in US.

## American Community Survey (ACS)
The American Community Survey is the survey conducted by the U.S. Census Bureau. The survey contains a variety of socio-economic variables for each county. Although there are different data tables from ACS, I will take 2021 five years estimate of DP02, DP03 and DP05.

These three tables cover the social, economic and demographic characteristics of each county in US. By referring my  previous research where I found the variables which have the highest correlations to the COVID mortality rate, I will use the same variables in this paper. That said, I will extract the data regarding the educational attainment from DP02, especially the proportion of people having at least a bachelor degree. In terms of economic perspective (DP03), I will take the proportion of people with private health 
```{r}
#| echo: false
#| message: false
#| label: tbl-covid
#| tbl-cap: The summary of COVID cases and deaths as of Election Day.
data %>% 
  group_by(winning_party) %>% 
  summarise(case = mean(cases),
            deaths = mean(deaths),
            inf = mean(infrate),
            mort = mean(mortrate))
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

# analysis_data <- read_csv(here::here("outputs/data/analysis_data.csv"))
# 
# analysis_data |> 
#   ggplot(aes(x = width, y = length)) +
#   geom_point(alpha = 0.8) +
#   theme_minimal() +
#   labs(x = "Wing width (mm)",
#        y = "Wing length (mm)")
```

Talk way more about it.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

@@

```{r}
# Counterfactual Analysis
adjusted_data <- data %>% 
  dplyr::select(state, county.x, fips, party, votes, total_votes, pct_vote, high_infrate)

>>>>

library(tidyverse)
library(here)
library(dplyr)
library(knitr)
library(ggplot2)
library(maps)
library(mapdata)
library(RColorBrewer)
library(gridExtra)
library(tools)
```

@@

## Center for Systems Science and Engineering at JHU

The Center for Systems Science and Engineering at JHU is the center at the Department of Civils and Engineering to collect the local, national and global multidimensional data including medicine, health care, disaster response etc. During the pandemic, they collected the US and global COVID cases and deaths and report them on their GitHub. Their data is summarized by daily reports, ranging from April 12, 2020 to March 9, 2023.

To illustrate the impact of COVID on the 2020 Election to the greatest extend, the COVID data used in this paper will be the daily report on November 3, 2020 which is the Election Day \[citation\]. This report contains the aggregated cases and deaths for each each country and counties in US.

## American Community Survey (ACS)

The American Community Survey is the survey conducted by the U.S. Census Bureau. The survey contains a variety of socio-economic variables for each county. Although there are different data tables from ACS, I will take 2021 five years estimate of DP02, DP03 and DP05.

These three tables cover the social, economic and demographic characteristics of each county in US. By referring my previous research where I found the variables which have the highest correlations to the COVID mortality rate, I will use the same variables in this paper. That said, I will extract the data regarding the educational attainment from DP02, especially the proportion of people having at least a bachelor degree. In terms of economic perspective, I will take the proportion of people with private health insurance and the mean household income. Lastly, regarding the demographic factors, I will use the total population, the proportions of children and individuals above 85. I will also take the White and Black people percentage as well.

## Data Cleaning

```{r}
#| echo: false
#| message: false
#| label: tbl-covid
#| tbl-cap: The summary of COVID cases and deaths as of Election Day.
data %>% 
  group_by(winning_party) %>% 
  summarise(case = mean(cases),
            deaths = mean(deaths),
            inf = mean(infrate),
            mort = mean(mortrate),
            income = mean(mean_household_income))
```

Combining the three data sources, @tbl-covid shows the COVID cases and deaths, as well as the infection and mortality rate for the county voting for the Democrat and Republican party, respectively. Surprisingly, the average case and deaths of COVID of the counties voting for the Democrat is significantly higher than than the Republican party. However, the values of infection and mortality rate does not show the same dramatic difference. It also emphasizes that the counties (states) voting for the Democrat are population intensive and the counties having more residence generally have higher mean household income level. We can observe this from the average income where the "Democratic" counties have about \$13k mean income higher than the "Republic" ones.

```{r}
#| echo: false
#| message: false
#| label: fig-boxplot
#| fig-cap: Summary of income levels for counties voting for Democratic and Republican

data  %>%
  ggplot(mapping = aes(x = winning_party, y = mean_household_income)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.04, width = 0.3, height = 0) +
  theme_minimal() +
  labs(x = "Voting Party",
       y = "Death Rate per 1000")
```

@fig-boxplot compares the distribution of county income levels for the two parties. In general, counties voting for the Democrat has a higher income levels than the Republican. In addition, we can observe that the counties are intensively concentrated around income levels \$50k to \$75k, compared to the Democrat where the counties are approximately uniform distributed at each income level. Furthermore, there is barely no counties having at least 150k voting for the Republican party. Both @tbl-covid and @fig-boxplot indicates that the Democrat is in favor of rich counties but poorer counties for Republican. Consequently, the richer counties (states) usually have more electoral votes than the poorer ones, this may provide insighes why Trump lost.

```{r}
#| echo: false
#| message: false

get_state_abbreviation <- function(state_names) {
  sapply(state_names, function(state_name) {
    state_data <- state.abb[match(state_name, state.name)]
    if (!is.na(state_data)) {
      return(state_data)
    } else {
      return(NA)  # changed from NULL to NA for consistency in a vector
    }
  })
}

county <- map_data("county")
state <- map_data("state")
county_clean <- county %>% 
  mutate(state = region,
         county = subregion,
         state = get_state_abbreviation(toTitleCase(state)),
         county = toTitleCase(county)) %>% 
  filter(region != "AK") %>% 
  select(-region, -subregion)

merged_data <- county_clean %>% left_join(republican_data, by = c("state", "county"))
```

```{r}
#| echo: false
#| message: false
#| fig.width: 10 
#| fig.height: 10
#| label: fig-map
#| fig-cap: The ratio of votes for the Republican and the infection rate per 100k in each county

p1 <- ggplot(data = merged_data, aes(x = long, y = lat, group = group)) + 
  geom_polygon(aes(fill = ratio_rd), color = "black") +
  scale_fill_distiller(palette = "RdBu", direction = -1, na.value = "grey") +
  labs(fill = 'Share of Republican') +
  coord_quickmap() +
  theme_void() +
  geom_path(data = state, aes(x = long, y = lat, group = group), color = "black", linewidth = 0.5)

# Second plot
p2 <- ggplot(data = merged_data, aes(x = long, y = lat, group = group)) + 
  geom_polygon(aes(fill = infrate), color = "black") +
  scale_fill_gradient(low = "white", high = "darkblue", na.value = "grey",
                      limits = c(0, 5000), oob = scales::squish) +
  labs(fill = 'Infection Rate per 100k') +
  coord_quickmap() +
  theme_void() +
  geom_path(data = state, aes(x = long, y = lat, group = group), color = "black", linewidth = 0.5)

# Arrange the plots in a 2x1 layout
grid.arrange(p1, p2, ncol = 1)
```

## Data Limitation

Although the merged data covers the voting patterns, socio-economic characteristics and COVID infection rate by county, some of the counties are missing. From the above two maps, the counties on the middle west shows a grey color, indicating that the data for these counties are absent. This absence of data may increase the bias and hence influence the accuracy of the data analyses.

# Methods

This paper aims to conduct the casual inference between the 2020 voting patterns and COVID, to find whether Trump can contribute his loss to the COVID. To find the casual effect in the observational study, I will implement the methods of propensity scores to find the change of votes for Trump due to the pandemic. Based on the results, I will also conduct a counterfactual analysis to see what would be happened if there was no COVID,

First and foremost, we need to define the treatment. The treatment in an experiment is something that will be artificially imposed to a certain groups of people to see the effect of interested variable with and without this treatment. Therefore the people with the treatment is called treatment group and the rest are called the control group. The settings of treatment and control group is to test whether the imposed factor really have impacts on the interested variable. However, since this is an observational study, it is unrealistic to impose a factor manually. In order to find the the casual effect of COVID on Trump's voting, I will choose the treatment to be whether a county has a high infection rate.

## Propensity Score

Propensity score is a technique that can help us to decide which county would be in the treatment group. That said, the propensity score in this study is the probability of a county having a high infection rate given a set of socioeconomic variables. By using the propensity score, it can ensure that the two counties with similar propensity score having the same distribution of the observed socio-economic variables such as income level even though they are in the different group. This is fantastic as we can have may pairs from treatment of control group which have similar covariates by fixing propensity scores. The key drawback that we can not estimate the casual effect from observational studies is addressed.

In addition to that, by finding the propensity scores, we can also reconstruct our assignment mechanism to a strongly ignorable one. An assignment mechanism is "strongly ignorable" if and only if, given the observed covariates, the potential outcomes are independent of the treatment assignment. This indicates that the change of voting for Trump is independent to our assignment that a county has high or low infection rate.

As the propensity score represents the probability of a county having high infection rate, therefore we need to find a logistic regression to predict this probability. On my previous paper, I already find the best model predicting the mortality rate of COVID in each US county. Even though that one was to predict the mortality rate, however, the mortality is correlated with infection rate. Therefore I will take the dependent variables in that model to be the covariates in predicting the propensity score for each county, which is:

```{=tex}
\begin{align*}
\frac{PS}{1-PS} &=  \beta_0 + \\
& + \beta_1 \times \text{prop\_higher\_education} \\
& + \beta_2 \times \text{IncomePctile} \\
& + \beta_3 \times \text{no\_insurance} \\
& + \beta_4 \times \text{private\_insurance} \\
& + \beta_5 \times \text{males} \\
& + \beta_6 \times \text{old\_85} \\
& + \beta_7 \times \text{white\_pct} \\
& + \beta_8 \times \text{black\_pct}
\end{align*}
```

Using the above model, we can therefore find the PS for each county. However, there are three different ways to use the propensity scores to find the treatment effect.

### Propensity Score Matching
The package needed to implement the matching is the `matching`. Using this package, we can match the observations in treatment and control group having the same or similar propensity scores. Then take only take the matched pairs to be the new data set and find the find the treatment effect between them. In addition, we can also check the balance between the treatment and control groups.

However, the most significant drawback is that the new data set can only contain the matched pairs. We may lose a huge amount of observations during this process. Our predictions may also not be accurate.

### Propensity Score Stratification
As introduced above that using PSM may cause the waste of data and results in bias. However, propensity score stratification will use the entire data set to avoid the waste. The way to do this is to split the data into different quantiles and fit a model for each subclass with the only dependent variable is the treatment variable. Then calculate the average casual effect for the subclasses.

Even though the PSS can use the entire data set, however, the choice of quantiles will be closely related to the predictions. If we have a large sample and want to split the data into many subclasses, then this method may be computational slow.

### Propensity Score Regression Adjustment
The last method is the most straightforward, meaning that we just fit a regression model with treatment variable and propensity score be the two dependent variables.

The above three methods have their unique advantages but the results may vary between different data sets. I will perform both the three methods and to compare which one would be the best.

## Counterfactural Analysis
By using the methodology of propensity score, we are able to access the casual effect of high infection rate on the votings for Trump. However, it can not tell us the election results and we can not make conclusions about the COVID and election. Therefore, I will conduct a counterfactural analysis, meaning that I will re-calculate the votes for Trump and Biden, to see if the final election results would be changed.

The way I will use to re-calculate the votes for the two candidates is to follow the official rule of election. By finding the difference of voting percentage, I will re-allocate the votings for the two parties and calculate the electoral votes in each state using the rule of winner-takes-all.

@@

```{r}
# Counterfactual Analysis
adjusted_data <- data %>% 
  dplyr::select(state, county, fips, party, votes, total_votes, pct_vote, high_infrate)
